Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-08-01 analysis/paper/paper.qmd
work on site description

# Kiwulan site and previous studies
This study focuses on archaeological evidence from Kiwulan, a large multi-component site where an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. Kiwulan was inhabited by the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in the historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with posts in situ. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. In addition to locally made pottery with geometric patterns, foreign goods such as porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials. Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area, mostly with the Basay people active in northern Taiwan.

# Previous research of

>>>>

# Archaeological context at Kiwulan
This study analyzes archaeological evidence from Kiwulan, a large multi-component site in Yilan, northeastern Taiwan. The upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) is the main focus [@Chen2007]. Kiwulan was inhabited by the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in the historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with wooden posts in situ. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. In addition to locally made pottery with geometric patterns, foreign goods such as porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials. Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area, mostly with the Basay people active in northern Taiwan. This reveals the continuous participation of Kiwulan in the regional trade network closely associated with northern Taiwan since 15th century [@]. Previous studies also indicates a possibly increased social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m), revealing numerous artifacts and features, including middens, burials, and post-holes with wooden posts in situ. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side, and burials located to the north of the residential area [@Chen2004]. Along with locally made pottery featuring geometric patterns, a variety of foreign goods—such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts—were found in both the residential area and burials. As there is no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@]. The use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

A wide range of animal bones and plant remains were found at Kiwulan, reflecting the diversity of local diet. The animals include deer, pigs, cows, birds, and fish, and plants mainly consisted of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale according to historical records [@]. Shellfish was also largely found, including bivalves such as clams and oysters, and univalves commonly from rivers and brackish water environments. Tools related to subsistence activities were found at Kiwulan, including iron arrowheads, iron sickles, and wooden paddles. The iron sickles indicate the cultivation and harvesting of crops, while the arrowheads suggest hunting activities targeting local wildlife such as deer and birds. The wooden paddle may have been used for fishing or other water-related activities, indicating a reliance on aquatic resources. The subsistence practices at Kiwulan should be primarily based on small scale rice farming, accompanied with hunting and fishing. Together, these findings suggest local residences engaged in a diverse range of subsistence activities, balancing agriculture with the exploitation of wild resources to meet their dietary needs.
